### 2012-05-23 doc/hmcsa_adjoint_vs_direct/direct_vs_adjoint.tex
Fixed a few typos in the direct vs adjoint tech note

%%---------------------------------------------------------------------------%%
\section{Introduction}
The Monte Carlo Synthetic Acceleration (MCSA) method developed by
Evans et. al. \cite{evans_2003} provides an acceleration to standard
stochastic linear solver methods that are bound by the central limit
theorem, instead giving exponential convergence. To do this, we must
first consider stochastic solutions for linear systems. For the
following linear system

@@

where k is the iteration index. In brief, in Eq. \ref{eq:mcsa_step1}
we compute an initial relaxation for $x$ with a Richardson iteration,
in Eq. \ref{eq:mcsa_step2} we compute the residual, in
Eq. \ref{eq:mcsa_step3} we compute the solution update via a
stochastic linear solve and finally we update the solution in
Eq. \ref{eq:mcsa_step4}. Of interest to this work is the stochastic
linear solution needed to compute the solution update in Eq.
\ref{eq:mcsa_step3} which we will choose to use either the direct or
adjoint method. It is important to note that the stochastic solver is
not computing $x$, rather it is computing an error term using the
residual that we will apply to $x$.

@@

Fig. \ref{fig:cpu_time} gives the CPU time needed for each case to
converge in seconds and Fig. \ref{fig:iterations} gives the number of
iterations needed for each case to converge. We see clearly in
Fig. \ref{fig:cpu_time} that the using the adjoint solver with MCSA
results in several orders of magnitude speedup over the direct solver
while the number of iterations required to converge is of a similar
scale. We expect this for several reasons. First, with an equivalent
number of histories specified for both solvers and a system of size
$N$, the direct solver will compute $N$ times the number of histories
random walks while the adjoint solver will only compute the number of
histories specified. Initially, this may not seem like a fair
comparison, however, we see from Fig. \ref{fig:cpu_time} that the
number of iterations requried to converge is approximately the
same. This is necessary in the direct method to ensure a contribution
from each state as the random walk sequence will only contribute to
the starting state. For the adjoint method, because the random
sequence contributes to the state in which it currently resides, fewer
histories are neccessary to compute contributions from all the nonzero
parts of the source.

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in]{AdjointDirectCPUTime.png}
  \caption{\sl CPU Time (s) to converge vs. Problem Size (N for an NxN
    problem). Both the adjoint and direct solvers are used with the
    five point and nine point stencils. A siginificant CPU time
    speedup is noted with the adjoint method due to the reduced number
    of histories.}
  \label{fig:cpu_time}
\end{figure}

>>>>

%%---------------------------------------------------------------------------%%
\section{Introduction}
The Monte Carlo Synthetic Acceleration (MCSA) method developed by
Evans et. al. \cite{evans_2003} provides an acceleration to standard
stochastic linear solver methods that are bound by the central limit
theorem by instead giving exponential convergence. However, a
component of the MCSA algorithm requires a linear stochastic solve to
still be performed. Both the direct and adjoint methods are available
to achieve this solution and will be studied here to assess if either
provides improved performance over the other.

We first consider an overview of stochastic solutions for linear
systems. For the following linear system

@@

where k is the iteration index. In brief, in Eq. \ref{eq:mcsa_step1}
we compute an initial relaxation for $x$ with a Richardson iteration,
in Eq. \ref{eq:mcsa_step2} we compute the residual, in
Eq. \ref{eq:mcsa_step3} we compute the solution update via a
stochastic linear solve and finally we update the solution in
Eq. \ref{eq:mcsa_step4}. Of interest to this work is the stochastic
linear solution needed to compute the solution update in Eq.
\ref{eq:mcsa_step3} for which we will choose to use either the direct
or adjoint method. It is important to note that the stochastic solver
is not computing $x$, rather it is computing an error term using the
residual that we will apply to $x$.

@@

Fig. \ref{fig:cpu_time} gives the CPU time needed for each case to
converge in seconds and Fig. \ref{fig:iterations} gives the number of
iterations needed for each case to converge. We see clearly in
Fig. \ref{fig:cpu_time} that the using the adjoint solver with MCSA
results in several orders of magnitude speedup over the direct solver
while the number of iterations required to converge is of a similar
scale. We expect this for several reasons. First, with an equivalent
number of histories specified for both solvers and a system of size
$N$, the direct solver will compute $N$ times the number of histories
random walks while the adjoint solver will only compute the number of
histories specified. Initially, this may not seem like a fair
comparison, however, we see from Fig. \ref{fig:cpu_time} that the
number of iterations requried to converge is approximately the
same. This is necessary in the direct method to ensure a contribution
from each state as the random walk sequence will only contribute to
the starting state. For the adjoint method, because the random
sequence contributes to the state in which it currently resides, fewer
histories are neccessary to compute contributions from all the nonzero
parts of the source. Therefore, although the total number of histories
per cycle is in fact larger for the direct method, the solution
computed at each iteration is no better than that computed by the
adjoint method.

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in]{AdjointDirectCPUTime.png}
  \caption{\sl CPU Time (s) to converge vs. Problem Size (N for an NxN
    problem). Both the adjoint and direct solvers are used with the
    five point and nine point stencils. A significant CPU time speedup
    is noted with the adjoint method due to the reduced number of
    histories.}
  \label{fig:cpu_time}
\end{figure}
